---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.path = "man/figures/README-", out.width = "100%")
```

# zigam

<!-- badges: start -->
[![R-CMD-check](https://github.com/AustralianAntarcticDivision/zigam/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/AustralianAntarcticDivision/zigam/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The **zigam** R package implements zero-inflated Poisson and Negative Binomial Generalized Additive Models from the **mgcv** package using the EM Algorithm.

## Installation

You can install the development version of zigam from [GitHub](https://github.com/) with:

```{r eval = FALSE}
# install.packages("remotes") ## if not already installed
remotes::install_github("AustralianAntarcticDivision/zigam")

```

## Example

A vignette (worked example) is available as part of the [package documentation](https://australianantarcticdivision.github.io/zigam) or can be view (after installation) using the code below:

```{r eval = FALSE}
browseVignettes("zigam")

```

